Case Study: scRNA-seq (Mouse data, scDD method)

### 2018-02-27 datasets/scrnaseq/scrnaseq-mouse-mast.Rmd
Organize sc analyses into one dataset/method per Rmd

---
title: "Case Study: scRNA-seq (Human data, MAST method)"
author: "Ayshwarya Subramanian and Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

Here we are interested in performing differential expression analyses to determine
which genes have different expression among the two biological groups in each comparison.
We also apply two different analysis methods: MAST and scDD. In this vignette, we will
explore the human dataset analysed with MAST.

@@

We first download the data in the data directory.`conquer` provides the processed datasets in convenient .rds files. First, we download the Human dataset:
GSE84465.

```{r, humandata-download}
download.file("http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE84465.rds",
  destfile=paste0(datdir, "GSE84465.rds"))

```

@@

We'll extract two conditions: Neoplastic and Regular cells, as
defined by their tsne clusters.

```{r}
datfile <- paste0(datdir, "human_GSE84465.rds")

@@

  # load MultiAssayExperiment
  humandata <- readRDS(paste0(datdir, "GSE84465.rds"))

  # subset by groups of interest
  humandata <- humandata2[, (colData(humandata)$characteristics_ch1.7 %in% 
                         c("neoplastic: Neoplastic", "neoplastic: Regular")) &
                         (colData(humandata)$characteristics_ch1.5 %in% 
                         c("tsne cluster: 11",  "tsne cluster: 8"))]
  cdata=colData(humandata)
  cdata$group <- cdata$characteristics_ch1.7

  # subset by gene expression TPM & count assays
  humandata <- experiments(humandata)[[1]]
  assays(humandata) <- assays(humandata)[c("TPM", "count")]
  colData(humandata) <- cdata

  # subset by genes not all zero
  humandata <- humandata[rowSums(assays(humandata)[["count"]]==0) < ncol(humandata),]

  # look at number of cells in each group
  table(colData(humandata)$group)

  # save SummarizedExperiment 
  saveRDS(humandata, datfile)
}else{
  humandata <- readRDS(datfile)
}
```

For the human dataset, there are a total of `r nrow(humandata2)` cells and `r ncol(humandata2)` genes.

@@

compute_MAST_sc(humandata)

# save results
saveRDS(humandata, datfile)
```

@@

humandata <- add_covariates_scrnaseq(humandata)

# save results
saveRDS(humandata, datfile)
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(humandata)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(humandata)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
```

>>>>

---
title: "Case Study: scRNA-seq (Mouse data, MAST method)"
author: "Ayshwarya Subramanian and Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

Here we are interested in performing differential expression analyses to determine
which genes have different expression among the two biological groups in each comparison.
We also apply two different analysis methods: MAST and scDD. In this vignette, we will
explore the mouse dataset analysed with MAST.

@@

We first download the data in the data directory.`conquer` provides the processed datasets in convenient .rds files. We download the Mouse data from GSE94383

```{r, mousedata-download}
download.file(url="http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE94383.rds",
  destfile=paste0(datdir,"GSE94383.rds"))
```

@@

We'll extract two conditions: No stimulation versus LPS at 150 min.

```{r}
datfile <- paste0(datdir, "mouse_GSE94383.rds")

@@

  # load MultiAssayExperiment
  mousedata=readRDS(paste0(datdir,"GSE94383.rds"))

  # subset by groups of interest
  mousedata <- mousedata[, (colData(mousedata)$characteristics_ch1.2 ==
                              "condition: No stimulation") |
                           (colData(mousedata)$characteristics_ch1.2 ==
                              "condition: LPS" &
                            colData(mousedata)$characteristics_ch1.1 %in% 
                              "time point: 150 min") ]
  cdata=colData(mousedata)
  cdata$group <- cdata$characteristics_ch1.2

  # subset by gene expression TPM & count assays
  mousedata <- experiments(mousedata)[[1]]
  assays(mousedata) <- assays(mousedata)[c("TPM", "count")]
  colData(mousedata) <- cdata

  # subset by genes not all zero
  mousedata <- mousedata2[rowSums(assays(mousedata)[["count"]]==0) < ncol(mousedata),]

  # look at number of cells in each group
  table(as.character(colData(mousedata2)$group))

  # save SummarizedExperiment 
  saveRDS(mousedata, datfile)
}else{
  mousedata <- readRDS(datfile) 
}
```

For the mouse dataset2, we have a total of `r ncol(mousedata2)` cells and 
`r nrow(mousedata2)` genes after filtering for non-detected genes.

@@

compute_MAST_sc(mousedata)

# save results
saveRDS(mousedata, datfile)
```

@@

mousedata <- add_covariates_scrnaseq(mousedata)

# save results
saveRDS(mousedata, datfile)
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
```

### 2018-02-27 datasets/scrnaseq/scrnaseq-mouse-scdd.Rmd
Organize sc analyses into one dataset/method per Rmd

---
title: "Case Study: scRNA-seq (Mouse data, MAST method)"
author: "Ayshwarya Subramanian and Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

Here we are interested in performing differential expression analyses to determine
which genes have different expression among the two biological groups in each comparison.
We also apply two different analysis methods: MAST and scDD. In this vignette, we will
explore the mouse dataset analysed with MAST.

@@

We perform differential expression testing using 
MAST which has a hurdle model to account for zeroes.

We apply MAST to the TPMs following the approach in Soneson et al.

```{r, MAST}
# function to add MAST p-values
compute_MAST_sc <- function(dat){
  # check if already computed
  if (! "mast_p" %in% colnames(rowData(dat))){
    sca <- FromMatrix(exprsArray = log2(assays(dat)[["TPM"]] + 1), 
                    cData = data.frame(wellKey=colnames(dat), colData(dat)))
    zlmdata <- zlm.SingleCellAssay(~group, sca)

    rowData(dat)$mast_p <- 
        lrTest(zlmdata, "group")[, "hurdle", "Pr(>Chisq)"]
  }else{
     message("MAST p-values already computed.")
  }
  return(dat)
}

compute_MAST_sc(mousedata)

# save results
saveRDS(mousedata, datfile)
```

@@

# save results
saveRDS(mousedata, datfile)
```

@@

For each covariate, we'll examine the plots for each of the two analysis methods.

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="meanExp") + 
  ggtitle("MAST, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="meanExp", maxy=25) 
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="mast_p", covariate="detection") + 
  ggtitle("MAST, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="mast_p", covariate="detection", maxy=25) 
```

>>>>

---
title: "Case Study: scRNA-seq (Mouse data, scDD method)"
author: "Ayshwarya Subramanian and Keegan Korthauer"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
        toc: true
        toc_float: true
        highlight: tango
        number_sections: true
---

@@

Here we are interested in performing differential expression analyses to determine
which genes have different expression among the two biological groups in each comparison.
We also apply two different analysis methods: MAST and scDD. In this vignette, we will
explore the mouse dataset analysed with scDD.

@@

We perform differential expression testing using scDD which implicitly models any heterogeneity in expression.

### scDD: Differences in distributions of gene expression in the 2 conditions

We apply scDD to normalized log-transformed counts. 
The p-value results are added to the `rowData` slot of the `SummarizedExperiment` 
data objects to ensure that gene metadata stays intact.

```{r, scdd}
# function to add scDD p-values
compute_scdd_sc <- function(dat){
  # check if already computed
  if (! "scdd_p" %in% colnames(rowData(dat))){
    sce <- SingleCellExperiment(assays=list(count=assays(dat)[["count"]]),
                                colData=data.frame(colData(dat)))
    # scran normalization
    libsize <- scran::computeSumFactors(assay(sce, "count"))
    normcounts(sce) <- GetNormalizedMat(assay(sce, "count"), libsize)

    rowData(dat)$scdd_p <- 
        results(scDD(sce, 
             condition = "group",
             prior_param = list(alpha=0.01, mu0=0, s0=0.01, a0=0.01, b0=0.01), 
             testZeroes = FALSE))$nonzero.pvalue
  }else{
     message("scDD p-values already computed.")
  }
  return(dat)
}

compute_scdd_sc(mousedata)

# save results
saveRDS(mousedata, datfile_m1)
```

@@

# save results
saveRDS(mousedata, datfile_m1)
```

@@

For each covariate, we'll examine the plots for each of the three analysis methods.

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="scdd_p", covariate="meanExp") + 
  ggtitle("scDD, Covariate 1: Mean Expression")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="meanExp", maxy=25) 
```

@@

```{r, fig.width=4.5, fig.height=3.5, message=FALSE}
rank_scatter(data.frame(rowData(mousedata)), 
             pvalue="scdd_p", covariate="detection") + 
  ggtitle("scDD, Covariate 2: Detection Rate")
```

```{r, fig.width=10, fig.height=3.2, message=FALSE}
strat_hist(data.frame(rowData(mousedata)), 
           pvalue="scdd_p", covariate="detection", maxy=25) 
```
